\emph{icepack}: a new glacier flow modeling package in Python, version 1.0

### 2021-04-02 reviews/response-3.tex

Added a reference to Ascher and Petzold.

\textcolor{blue}{56, equation (1) and nearby, AB:  Two concerns.  First, it is later acknowledged that this is not really an advection equation (lines 376--377), so one does not need to call it that here either.  The SIA is not some weird alternative universe of weak-willed modelers, it is what *all models should produce* in the large.  That model, the only clearly-understood coupled model, makes equation (1) a diffusion, as noted.  Surely calling this a "transport equation" or even "thickness transport equation" is adequate.  One then points-out that $q=hu$ is one way to parameterize flux, and that $u$ comes from a coupled equation \emph{driven by $\nabla s$} (even in the Stokes case).  It might be acknowledged that (1) does not have a PDE "type" in the classical sense.  (It is a DAE, after all.)  Second, equation (1) holds with what boundary conditions?  This manuscript maintains the tradition of pretending not to notice.  To quote [Schoof \& Hewitt 2013], "A sometimes weakly perceived point in glaciology is that the model above is in fact a free-boundary problem ...", and *this applies to (1) regardless of the stress balance model*.  Is this paper just going to pretend boundary conditions for the main, and first, equation don't exist?  (Or pretend that all glaciers end in cliffs of sufficient height so that the singular change of coordinates (16) is no problem?)  This manuscript could even cite existing wellposedness literature for (1), especially [Calvo et al 2003] and [Jouvet \& Bueler 2012], which confirm the meaningfulness of the free-boundary problem formulation and the resultant simultaneous Dirichlet (h=0) and Neumann (q=0) boundary conditions at grounded margins.  And then say what Icepack will when solving (1).}

We removed the statement that this is an advection equation and added the following sentences: ``This problem has the apparent form of a conservative advection equation.
The velocity $u$ is coupled to the thickness and surface slope, however, in such a way that the whole problem is not hyperbolic and in many cases does not have a PDE `type' in the usual sense.''

@@

\textcolor{blue}{298--299:  Suggested replacement for the mushy phrase: "describing realistic glacier flows, but it is not sufficient by itself" --> "parameterizing fluidity".}

Done.

\textcolor{blue}{300--301, C:  I seriously doubt that more than one user in 100 will want to substitute their own melt-fraction dependence, because no one has the data for it.  So: "likely" --> "possible".  (This is a clear case where good defaults are more important than modularity.)}

Changed to ``Some users of this module may want to substitute in their own parameterization for melt fraction dependence.''

@@

\textcolor{blue}{297--305, C:  This paragraph reads, to me, like an argument that expert knowledge *will* be required for the Icepack user.  ("users must calculate the fluidity field themselves")  That is, unless Icepack includes both defaults and higher-level simulation drivers which allow the novice user to never know about these possibilities.  More broadly, what does Icepack design offer, when it comes to avoiding explicit choices of all submodels, on the way to a first scientifically-useful model for the beginner?}

What you're describing is knowledge of glaciology, not scientific computing.
We have stated explicitly in the first sentence of section 6 that we aim to reduce the need for users to be experts in the latter but not the former.
If we were to define how some fields are coupled, we would be faced with a difficult design choice of where this is or is not appropriate.
Including the effect of temperature on rheology seems fairly obvious.
Some modeling papers also account for much higher fluidity for ice that accumulated during the last glacial period because of the greater dust content.
Should it also be the responsibility of the package to include this effect?
We did not feel that, at this stage of the development, we could also implement high-level simulation drivers that hide these details while also satisfying our other design goals.

@@

\textcolor{blue}{318--320:  Mushy sentences.  Needed?}

We wanted to emphasize that this model is very ad hoc and that we're not passing it off as more accurate than it is.
These sentences have been removed.

\textcolor{blue}{345, C:  The sentence "The key classes that users interact with are flow models and solvers" is right at the heart of my concern C.  Indeed, after developing PISM with much (developer group) concern about flow models and solvers, we spent the next 10 years talking to users about their data formats, parameter-study schema, and surface process models (supra, sub, and calving front).  Barely had time to work on flow models and solvers ... until I quit day-to-day PISM development.  The by-far largest population of scientists care about how flowing ice interacts with their climates.  They use ice flow models on the assumption that ice flow modelers know how to model ice flow!  This "key classes" sentence is not describing users, it is describing co-developers.}

Again I think this seems to be a disagreement about what the software package should do and what the users should be responsible for.
Icepack lacks any of the high-level simulation drivers that PISM has.
Giving users a relatively lower level interface to the physics solvers was a concession to the fact that we can't predict what they will want to do with an ice flow model by way of coupling to climate, oceans, solid earth, etc.
Obviously you and the developers of PISM have had far more experience.
We may well have misjudged what glaciologists do and don't find difficult about modeling and what type of interface best serves the needs of the most people.

@@

\textcolor{blue}{721:  "phyics"}

Fixed.

\textcolor{blue}{724:  Having reviewed quite a few ice sheet modeling papers, every single one claimed something about its usability.  Less mush, more answers please.}

Changed to ``Relatively few works in the computational science literature draw directly from relevant work in HCI when discussing usability; see for example Hannay et al. 2009, Harris et al. 2020.''

@@

\textcolor{blue}{725:  I'm on board with (2), and wish for more careful analysis of Icepack's contribution here, but "quantifying" in (1) is not what you are doing in this manuscript.  (Nor, probably, do you have the ability to do it.)}

Changed from ``quantifying'' to ``evaluating''.
We're making an unsubstantiated claim that usability is indeed a rate-limiting factor.

### 2021-04-02 icepack.tex
More responses to reviewer 3

The two main components of a glacier flow model are a \emph{diagnostic} and a \emph{prognostic} equation.
The diagnostic equation prescribes the ice velocity through a time-independent, nonlinear, elliptic partial differential equation (PDE).
The prognostic equation prescribes how the ice thickness evolves through conservation of mass.
Mathematically, these two coupled PDEs can be thought of as a differential-algebraic equation.

@@

The \emph{prognostic model} or \emph{mass transport equation} (the two terms are synonymous) describes how the ice thickness changes in time.
The prognostic model is a conservative advection equation:
\begin{equation}
    \frac{\partial h}{\partial t} + \nabla\cdot hu = \dot a_s - \dot a_b,
\end{equation}
where $\dot a_s$ and $\dot a_b$ are the surface and basal mass balance.
Icepack represents the thickness using continuous, piecewise polynomial basis functions in each cell of the mesh.
We have not yet implemented a formulation that works with discontinuous basis functions, but this extension is completely feasible within our framework.

@@

We have also paid special attention to how we can design this software package to be most usable for its intended audience.
Usability is not often discussed in the literature on computational science.
We believe that this is because of two difficulties: (1) quantifying the degree to which usability is a rate-limiting factor and (2) concretely assessing what features make software tools more or less usable.
By contrast, measuring computational performance is much more feasible although still fraught with difficulties of its own.
(This is not to say that performance optimization is easy by any means.)
In working with several graduate students and postdoctoral researchers in glaciology, we have observed that usability is a substantial bottleneck for researchers at these career stages.
For this reason, we have chosen to focus explicitly on usability by drawing on the research literature in HCI.
We argue that the same design features that enhance usability for relative novices to the subject area will also enhance the productivity of expert users.

>>>>

The two main components of a glacier flow model are a \emph{diagnostic} and a \emph{prognostic} equation.
The diagnostic equation prescribes the ice velocity through a time-independent, nonlinear, elliptic partial differential equation (PDE).
The prognostic equation prescribes how the ice thickness evolves through conservation of mass.
Mathematically, these two coupled PDEs can be thought of as a differential-algebraic equation; see \citet{ascher1998computer}.

@@

The \emph{prognostic model} or \emph{mass transport equation} (the two terms are synonymous) describes how the ice thickness changes in time.
The prognostic model is:
\begin{equation}
    \frac{\partial h}{\partial t} + \nabla\cdot hu = \dot a_s - \dot a_b,
\end{equation}
where $\dot a_s$ and $\dot a_b$ are the surface and basal mass balance.
This problem has the apparent form of a conservative advection equation.
The velocity $u$ is coupled to the thickness and surface slope, however, in such a way that the whole problem is not hyperbolic and in many cases does not have a PDE ``type'' in the usual sense.

Icepack represents the thickness using continuous, piecewise polynomial basis functions in each cell of the mesh.
We have not yet implemented a formulation that works with discontinuous basis functions, but this extension is completely feasible within our framework.

@@

We have also paid special attention to how we can design this software package to be most usable for its intended audience.
Relatively few works in the computational science literature draw directly from relevant work in HCI when discussing usability; see for example \citet{hannay2009scientists, harris2020array}.
We believe that this is because of two difficulties: (1) evaluating the degree to which usability is a rate-limiting factor and (2) concretely assessing what features make software tools more or less usable.
By contrast, measuring computational performance is much more feasible although still fraught with difficulties of its own.
(This is not to say that performance optimization is easy by any means.)
In working with several graduate students and postdoctoral researchers in glaciology, we have observed that usability is a substantial bottleneck for researchers at these career stages.
For this reason, we have chosen to focus explicitly on usability by drawing on the research literature in HCI.
We argue that the same design features that enhance usability for relative novices to the subject area will also enhance the productivity of expert users.
